---
title: "HW5"
author: "Team_9"
date: "2023-02-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 4

## Team 9:

-   Charlie Madison
-   Hrishi Mysore Harishkumar
-   Michelle Li
-   Qizhuang Huang
-   Shaun Pfister
-   Rodolfo Lerma

## Description

The Publication data in the ISLR2 R package contains information about the time to publication for the results of 244 clinical trials funded by the National Heart, Lung, and Blood Institute. Take some time to read more about this dataset in chapter 11.5.4 of ISL. You can also type ?Publication in R for more information after loading the dataset.

Data frame with 244 observations, each representing a clinical trial, and 9 variables:

-   `posres`: Did the trial produce a positive (significant) result? 1=Yes, 0=No.
-   `multi`: Did the trial involve multiple centers? 1=Yes, 0=No.
-   `clinend`: Did the trial focus on a clinical endpoint? 1=Yes, 0=No.
-   `mech`: Funding mechanism within National Institute of Health: a qualitative variable.
-   `sampsize`: Sample size for the trial.
-   `budget`: Budget of the trial, in millions of dollars.
-   `impact`: Impact of the trial; this is related to the number of publications.
-   `time`: Time to publication, in months.
-   `status`: Whether or not the trial was published at time: 1=Published, 0=Not yet published.

## Question 1:

**Load the Publication dataset in R.**

```{r, message = FALSE}
library(ISLR2)
library(survival)
```

```{r}
data(Publication)
head(Publication)
```

## Question 2:

**Calculate for how many clinical trials the time associated with the event of interest (i.e., the time of publication) is observed and calculate for how many clinical trials the time of the event of interest is censored.**

```{r}
# Calculate the number of observed and censored events
table(Publication$status)
```

We can see that out of the 244 trials 156 ended up with being published (and therefore we have the time of Publication) and at the same time we can notice that the other 88 cases are Censored as the publications has not yet happened.

## Question 3:

**Produce and plot the `Kaplan-Meier` estimator for the time to publication of all the clinical trials in the dataset. Include 99% pointwise confidence bands in the plot.**

```{r}
# Fit a survival model
km <- survfit(Surv(time, status) ~ 1, data = Publication)

#Plot
plot(
    km,
    conf.int = 0.99,
    col = "blue",
    xlab = "Months",
    ylab = "Estimated Survival Function"
)

grid(nx = NULL, ny = NULL,
     lty = 2, col = "gray", lwd = 1)

title(main = "Kaplan-Meier Estimator")
```

## Question 4:

**Produce and plot the `Kaplan-Meier` estimator for the time to publication of the clinical trials for the two subgroups corresponding to the `posres` variable (i.e., for the group of clinical trials that resulted in positive findings and for the group of clinical trials that did not result in positive findings).**

```{r}
# Fit a survival model
km_posres <- survfit(Surv(time, status) ~ posres, data = Publication)

# Plot
plot(
  km_posres,
  conf.int = 0.95,
  col = c("purple", "violet"),
  xlab = "Months",
  ylab = "Estimated Survival Function"
)

# Convert posres to factor and specify levels for the legend
Publication$posres <- as.factor(Publication$posres)
levels(Publication$posres) <- c("NO", "YES")

# Add legend, grid, and title
legend(
  4, 0.2,
  levels(Publication$posres),
  col = c("purple", "violet"),
  lty = 2,
  box.lty = 0
)

grid(nx = NULL, ny = NULL,
     lty = 2, col = "gray", lwd = 1)

title(main = "Kaplan-Meier Estimator (Posres)")
```

**Then, use the log-rank test to test the null hypothesis that the time to publication is not associated with whether or not the clinical trial resulted in a positive finding (`posres`). State in English the result of the log-rank test.**

```{r}
posres_logrank <- survdiff(Surv(time, status) ~ posres, data = Publication)
posres_logrank
```

## Question 5:

**Fit a Cox proportional hazards model to these data using the following predictors:**

-   `posres`
-   `multi`
-   `clinend`
-   `budget`

**Also produce the model summary with the `summary` function.**

```{r}
cox_model <- coxph(Surv(time, status) ~ posres + multi + clinend + budget, data = Publication)
summary(cox_model)
```

## Question 6:

**Do the global likelihood ratio, Wald, and score test suggest that the model is better than a model that does not use any predictor?**

Given the very small P Value (2e-16) we concluded that yes, these parameters suggest the model is better than a model with no predictors.

## Question 7:

**In English, interpret the estimated effect of each predictor on the hazard function corresponding to the time to publication.**

-   `posresYES`: Publication with a positive result have an estimated hazard ratio of 1.705 (95% CI: 1.2024 to 2.418) compared to those with negative results. This means that the hazard of publishing for the group with positive results is 1.705 times higher than the hazard of publishing for the group with negative results, after adjusting for the other covariates in the model.
-   `multi`: Multi-centre studies have an estimated hazard ratio of 1.884 (95% CI: 1.2054 to 2.945) compared to single-centre studies. This means that the hazard of publishing for multi-centre studies is 1.884 times higher than the hazard of publishing for single-centre studies, after adjusting for the other covariates in the model.
-   `clinend`: Clinical endpoint studies have an estimated hazard ratio of 5.163 (95% CI: 3.2172 to 8.287) compared to studies without clinical endpoints. This means that the hazard of publishing for studies with clinical endpoints is 5.163 times higher than the hazard of publishing for studies without clinical endpoints, after adjusting for the other covariates in the model.
-   `budget`: The effect of budget on the hazard of publication is not statistically significant at the 5% level (p-value = 0.20477). This means that we cannot conclude that there is a significant effect of budget on the hazard of publishing, after adjusting for the other covariates in the model. The estimated hazard ratio is 1.002 (95% CI: 0.9988 to 1.006), which suggests that there is practically no effect of budget on the hazard of publishing.

```{r}
cox_model_baseline <- basehaz(cox_model, centered = FALSE)
plot(
    cox_model_baseline$time,
    cox_model_baseline$hazard,
    type = "b",
    xlab = "Months",
    ylab = "Cumulative hazard"
)

grid(nx = NULL, ny = NULL,
     lty = 2, col = "gray", lwd = 1)

title(main = "Cumulative Baseline Hazard Function")
```

## Question 8:

**First, use the above commands to plot the estimated baseline survival function S0 for the time to publication (this is obtained by setting all predictors to 0 in newdata). Then, use the above commands to plot**

-   the estimated survival function for the time to publication of a new clinical trial with predictor values:
    -   posres: 0
    -   multi: 0
    -   clinend: 1
    -   budget: 8.5
-   the estimated survival function for the time to publication of a new clinical trial with predictor values
    -   posres: 0
    -   multi: 0
    -   clinend: 0
    -   budget: 1.3

**Based on the two estimated survival functions, which of these two clinical trials do you think is more likely to be published sooner? Explain.**


```{r}
# Create a dataframe with the predictor values for a new clinical trial
new_data <- data.frame(posres = 0, multi = 0, clinend = 1, budget = 8.5)

# Calculate the estimated survival function for the baseline (all predictors set to 0)
est_surv_base <- survfit(cox_model, newdata = data.frame(posres = 0, multi = 0, clinend = 0, budget = 0))

# Plot the estimated baseline survival function
plot(est_surv_base, xlab = "Time to Publication", ylab = "Survival Probability", main = "Estimated Baseline Survival Function")

# Calculate the estimated survival function for the new clinical trial
est_surv_new <- survfit(cox_model, newdata = new_data)

# Plot the estimated survival function for the new clinical trial
plot(est_surv_new, xlab = "Time to Publication", ylab = "Survival Probability", main = "Estimated Survival Function for New Clinical Trial")
```

